---
title: "PSTAT174 Final Project"
author: "Aaron Lee (3410388)"
date: "2023-05-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Energy Generation Data

```{r}
electricity.csv <- read.table("electricity_data.csv", sep = ",", header = FALSE, skip = 1, nrows = 255)
electricity <- ts(electricity.csv[, 2], start = c(2001, 1, 1), frequency=12)

electricity1 = electricity[c(1: 243)]/100000
electricity1_test = electricity[c(244: 255)]/100000
```

## Plotting the original data

```{r}
ts.plot(electricity1, main="Monthly Electricity Generation in all sector of US", ylab=expression(X[t]))
ele_fit <- lm(electricity1 ~ as.numeric(1:length(electricity1))); abline(ele_fit, col="red")
abline(h=mean(electricity1), col="blue")
```

* Not stationary; we can see that there is an upper trend; seasonality; No constant variance.

```{r}
hist(electricity1, main="Monthly Electricity Generation in all sector of US", xlab="Monthly electricity generation")
acf(electricity1)
```

```{r}
library(MASS)
t <- 1:length(electricity1)
bcTransform <- boxcox(electricity1 ~ t, plotit=TRUE)
bcTransform$x[which(bcTransform$y == max(bcTransform$y))]
```

```{r}
lambda = bcTransform$x[which(bcTransform$y == max(bcTransform$y))]
electricity1.bc = (1/lambda)*(electricity1^lambda-1)
electricity1.log = log(electricity1)
electricity1.sqrt = sqrt(electricity1)

op <- par(mfrow = c(2,2))
ts.plot(electricity1,main = "Original data",ylab = expression(X[t])) 
ts.plot(electricity1.bc,main = "Box-Cox tranformed data", ylab = expression(Y[t]))
ts.plot(electricity1.log, main = "Log Transform")
ts.plot(electricity1.sqrt, main = "Square Root Transform")
par(op)
```

```{r}
hist(electricity1.bc)
```

* Through the histogram, we choose box-cox transformation.

### Produce decomposition of Box-Cox $U_t$

```{r}
library(ggplot2)
#install.packages('ggfortify')
library(ggfortify)

y <- ts(as.ts(electricity1.bc), frequency = 12)
decomp <- decompose(y)
plot(decomp)
```

```{r}
# Calculate the sample variance and plot the acf/pacf
var(electricity1)
var(electricity1.bc)
```
The variance increases after the transformation.

```{r}
op = par(mfrow = c(1,2))
acf(electricity1.bc,lag.max = 40,main = "")
pacf(electricity1.bc,lag.max = 40,main = "")
title("Box-Cox Transformed Time Series", line = -1, outer=TRUE)
par(op)
```

```{r}
# Difference at lag  = 1 to remove trend component
y1 = diff(electricity1.bc, 1)
plot.ts(y1,main = "De-trended Time Series",ylab = expression(nabla~Y[t]))
abline(h = 0,lty = 2)
var(y1) # smaller that 7.342619e-5
```

```{r}
# Difference at lag  = 12 (cycle determined by the ACF) to remove seasonal component
y12 = diff(y1, 12)
plot.ts(y12, main = "De-trended/seasonalized Time Series", ylab = expression(nabla^{12}~nabla~Y[t]))
abline(h = 0,lty = 2)
var(y12) # smaller than 7.342619e-5 and 6.753134e-5
```

```{r}
acf(y12,lag.max = 40,main = "")
title("ACF: First and Seasonally Differenced Time Series", line = -1, outer = TRUE)
pacf(y12,lag.max = 40,main = "")
title("PACF: First and Seasonally Differenced Time Series", line = -1, outer = TRUE)
hist(y12, col="light blue", xlab="", main="histogram; ln(U_t) differenced at lags 12 & 1")
```

```{r}
# Compare histograms of Box-Cox (Ut) to the normal curve, really similar.
hist(y12, density=20,breaks=20, col="blue", xlab="", prob=TRUE)
m<-mean(y12)
std<- sqrt(var(y12))
curve( dnorm(x,m,std), add=TRUE )
```

#### **Modeling the seasonal part** (P, D, Q): For this part, focus on the seasonal lags h = 1s, 2s, etc.

* We applied one seasonal differencing so D = 1 at lag s = 12.
* The ACF shows a strong peak at h = 1s and smaller peaks appearing at h = 2s.  A good choice for the MA part could be Q = 1 or Q = 2.
* The PACF shows there is a peak at h = 1s.  A good choice for the AR part could be P = 1.

#### **Modeling the non-seasonal part** (p , d, q): In this case focus on the within season lags, h = 1,. . . ,11.

* We applied one differencing to remove the trend: d = 1.
* A good choice for the MA part could be q = 0 or q = 1 respectively.
* A good choice for the AR part could be p = 2

#### Also, the model might be MA(33); SARIMA(2,1,0)(1,1,1)[12]; SARIMA(2,1,1)(1,1,1)[12]; SARIMA(2,1,0)(1,1,2)[12]; SARIMA(2,1,1)(1,1,2)[12]

### Trying Models:

#### SMA models tried: Q=1, 2, q=0,1. Model producing the lowest AICc:
```{r}
library(astsa)
library(MuMIn)
arima(electricity1.bc, order = c(0,1,1), seasonal = list(order = c(0,1,2), period = 12), method="ML")
# Calculating AICc
AICc(arima(electricity1.bc, order = c(0,1,1), seasonal = list(order = c(0,1,2), period = 12), method="ML"))
```

```{r}
arima(electricity1.bc, order = c(0,1,0), seasonal = list(order = c(0,1,2), period = 12), method="ML")
AICc(arima(electricity1.bc, order = c(0,1,0), seasonal = list(order = c(0,1,2), period = 12), method="ML"))
```

```{r}
arima(electricity1.bc, order = c(0,1,1), seasonal = list(order = c(0,1,1), period = 12), method="ML")
AICc(arima(electricity1.bc, order = c(0,1,1), seasonal = list(order = c(0,1,1), period = 12), method="ML"))
```

```{r}
arima(electricity1.bc, order = c(0,1,0), seasonal = list(order = c(0,1,1), period = 12), method="ML")
AICc(arima(electricity1.bc, order = c(0,1,0), seasonal = list(order = c(0,1,2), period = 12), method="ML"))
```

### SAR
```{r}
arima(electricity1.bc, order = c(2,1,0), seasonal = list(order = c(1,1,0), period = 12), method="ML")
AICc(arima(electricity1.bc, order = c(2,1,0), seasonal = list(order = c(1,1,0), period = 12), method="ML"))
```

### SARIMA(2,1,1)(1,1,2)_s=12
```{r}
arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), method="ML")
AICc(arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), method="ML"))
```

### Best fit model
```{r}
arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), fixed = c(NA,0,NA,NA,NA,NA),method="ML")
AICc(arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), fixed = c(NA,0,NA,NA,NA,NA),method="ML"))
```

### MA(33) AICc is not smaller than -2102.679
```{r}
arima(electricity1.bc, order = c(0,0,33), seasonal = list(order = c(0,0,0), period = 12),method="ML")
AICc(arima(electricity1.bc, order = c(0,0,33), seasonal = list(order = c(0,0,0), period = 12),method="ML"))
```

```{r}
arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,1), period = 12),method="ML")
AICc(arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,1), period = 12),method="ML"))
```

### second less AICc
```{r}
arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,1), period = 12), fixed = c(NA,0,NA,NA,NA),method="ML")
AICc(arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,1), period = 12), fixed = c(NA,0,NA,NA,NA),method="ML"))
```
* not invertible, because 1.0001 is bigger than 1


### Diagnostic checking
```{r}
#fit <- arima(electricity1.bc, order=c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), method="ML")
fit <- arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), fixed = c(NA,0,NA,NA,NA,NA),method="ML")
res <- residuals(fit)
hist(res,density=20,breaks=20, col="blue", xlab="", prob=TRUE)
m <- mean(res)
std <- sqrt(var(res))
curve( dnorm(x,m,std), add=TRUE )
plot.ts(res)
fitt <- lm(res~as.numeric(1:length(res))); abline(fitt, col="red") 
abline(h=mean(res), col="blue")
```

```{r}
qqnorm(res,main= "Normal Q-Q Plot for Model SARIMA(2,1,1)(1,1,2)_[12]")
qqline(res,col="blue")
```

```{r}
acf(res, lag.max=40)
pacf(res, lag.max=40)
```

```{r}
fit.i <- sarima(xdata=electricity1, p=2, d=1, q=1, P=1, D=1, Q=2, S=12)
```


```{r}
# p-value should be bigger that 0.05
shapiro.test(res) # p-value should be bigger that 0.05
Box.test(res, lag = 16, type = c("Box-Pierce"), fitdf = 3)
Box.test(res, lag = 16, type = c("Ljung-Box"), fitdf = 3)
Box.test(res^2, lag = 16, type = c("Ljung-Box"), fitdf = 0)
```
All p-value is larger than 0.05.

```{r}
acf(res^2, lag.max=40) # do not need this
ar(res, aic = TRUE,  order.max = NULL, method = c("yule-walker"))
```

Fitted residual to AR(0), White noise
Pass Diagnostic checking. 
Ready to be used for forecasting.

```{r}
library(forecast)
pred.tr <- sarima.for(electricity1.bc,n.ahead = 12,p=2,d=1,q=1,P=1,D=1,Q=2, S=12)
#sarima.for(electricity1.bc,n.ahead = 12,p=2,d=1,q=1,P=1,D=1,Q=2, S=12)
points(length(electricity1) + 1:length(electricity1_test),pred.tr$pred, col="blue",pch = 19)

```

```{r}
#pred.orig <- InvBoxCox(pred.tr$pred, lambda)
#sarima.for(electricity1,n.ahead = 12,p=2,d=1,q=1,P=1,D=1,Q=2, S=12)
#ts.plot(electricity1, xlim=c(1,length(electricity1)+12))
#points(length(electricity1) + 1:length(electricity1_test),electricity1_test, col="blue",pch = 19)
```


```{r}
# arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), fixed = c(NA,0,NA,NA,NA,NA),method="ML")

# Forecasting using model A:
fit.A <- arima(electricity1.bc, order = c(2,1,1), seasonal = list(order = c(1,1,2), period = 12), fixed = c(NA,0,NA,NA,NA,NA),method="ML")
forecast(fit.A) 

# To produce graph with 12 forecasts on transformed data:
pred.tr1 <- predict(fit.A, n.ahead = 12)
U.tr = pred.tr1$pred + 2*pred.tr1$se # upper bound of the prediction interval
L.tr = pred.tr1$pred - 2*pred.tr1$se # lower bound
plot.ts(electricity1.bc, xlim=c(1,length(electricity1.bc)+12), ylim = c(min(electricity1.bc), max(U.tr)))
lines(U.tr, col="blue",lty="dashed")
lines(L.tr, col="blue",lty="dashed")
points((length(electricity1.bc)+1):(length(electricity1.bc)+12), pred.tr1$pred, col="red",pch = 19)
```
```{r}
# To produce graph with forecasts on original data:
pred.orig <- InvBoxCox(pred.tr1$pred, lambda)
U= InvBoxCox(U.tr, lambda)
L= InvBoxCox(L.tr, lambda)
plot.ts(electricity1, xlim=c(1,length(electricity1)+12), ylim = c(min(electricity1),max(U)))
lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")
points((length(electricity1)+1):(length(electricity1)+12), pred.orig, col="red",pch = 19)

# To zoom the graph, starting from entry 200
ts.plot(electricity1, xlim = c(200,length(electricity1)+12), ylim = c(min(electricity1),max(U)))
lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")
points((length(electricity1)+1):(length(electricity1)+12), pred.orig, col="red")
```
```{r}
# To plot zoomed forecasts and true values (in electricity):
electricity_true <- electricity[1:255]/100000
plot.ts(electricity_true, xlim = c(200,length(electricity1)+12), ylim = c(2.7,max(U)), col="red")
lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")
points((length(electricity1)+1):(length(electricity1)+12), pred.orig, col="green")
points((length(electricity1)+1):(length(electricity1)+12), pred.orig, col="black")
```



